Unsupervised embeddings of organizations capture latent structure of scientific mobility

### 2020-04-20 paper/current/main.tex
additional language stressing linguistic properties of embeddings

As scientist's careers progress, they move between organizations or pick up additional affiliations, forming co-affiliation or \textit{mobility trajectories} that can be derived from their publication histories~\autocite{robinson2019mobility, sugimoto2017mostimpact}.
By treating geo-located organizations analogously to words and trajectories analogously to sentences, we learn vector-space embedding of organizations using a neural embedding method (Fig.~\ref{fig:methods:mobility_sentence}).
In the resulting embedding, each organization corresponds to a vector, and the \textit{cosine distance} between vectors represents the distance in the landscape of scientific mobility.

@@

Where the \textit{expected flux}, $\hat{T_{ij}}$, or expected number of authors co-affiliated with a pair organizations during the time period is proportional to the organization's sizes, $m_{i}$ and $m_{j}$, and decays as a function of their distance, $f(r_{ij})$.
We consider two distances, the geographic distance between the organizations as a baseline, and the \textit{embedding distance} defined as the cosine distance between the organization's vector in the trained embedding space.
We argue that the performance of a distance measure can be observed from their ability to explain expected mobility between organizations and to make predictions using the gravity model.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

Given that embeddings capture effective distance between organizations, we can now examine the global structure of scientific mobility.
Because the embeddings are defined in a high-dimensional space, we apply a widely-used dimensionality reduction technique (UMAP~\autocite{mcinnes2018umap}) to visualize the overall structure (Fig.~\ref{fig:projection}a).
Globally, the effect of geography is conspicuous; organizations tend to form clusters based on their national affiliations and national clusters tend to be near their geographic neighbors. 
At the same time, the embedding space also reflects a mix of geographic, historic, cultural, and linguistic relationships between regions much more clearly than altenrative network visualizations (Fig.~\ref{fig:supp:network_vis}).

@@

The results demonstrating embeddings' ability to capture various features governing mobility raise the question of whether they can also capture semantic axes analogously to word embeddings. 
For instance, does the mobility embedding encode information about the prestige or universities?
In addition to encoding national characteristics, the embedding also encodes domestic geography and latent hierarchy.
Leveraging the unique semantic properties of the organization embedding, we employ the SemAxis approach~\autocite{an2018semaxis} and define two semantic axes.
The first is the California--Massachusetts axis defined by the mean vector of universities in California on one end, and in Massachusetts on the other. 
The second axis is the low- to high- prestige axis;
the elite end of the axis is defined as the top 20 U.S. universities according to the Times Ranking of World Universities, whereas the non-elite axis is defined as bottom-ranked and geographically matched (by U.S. census region) set of universities.
A university projected onto the axis corresponds to their nearness to the mean vector defining each pole (Fig.~\ref{fig:semaxis}a).

Generally, universities in Western states, such as Arizona (Fig.~\ref{fig:supp:semaxis_states}b) are nearer to the California end of the axis, whereas states in the Northeast such as New York (Fig.~\ref{fig:supp:semaxis_states}f) and Connecticut (Fig.~\ref{fig:supp:semaxis_states}e) are nearer the Massachusetts end.
Universities geographically distant from both California and Massahcussets are less aligned with either, such as Texas (Fig.~\ref{fig:supp:semaxis_states}g)s and Florida (Fig.~\ref{fig:supp:semaxis_states}i).

Universities hiring and mobility networks are known to exhbit latent pretige hierarchies ~\autocite{deville2014career, clauset2015hierarchy} that should be encoded in the embedding space.
Accordingly, prestigious universities such as Stanford, Colombia, MIT, Harvard, and NYU, group towards the top of the axis.
Universities at the bottom of this axis tend to be regional universities with lower national profiles.
By considering mobility between all organization types, organization embeddings offer a new means of reasoning about the prestige of organizations typically excluded from rankings such as government organizatioss (Fig.~\ref{fig:supp:semaxis_sectors}a), research institutes (Fig.~\ref{fig:supp:semaxis_sectors}b), and teaching colleges (Fig.~\ref{fig:supp:semaxis_sectors}c)

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_fig.pdf}
	\caption{
		\textbf{Embedding captures latent geography and prestige hierarchy.}
		\textbf{a.} illustrates SemAxis projection along two axes, comparing California to Massachusetts universities (left to right), and between the top 20 and geographically-matched bottom 20 universities ranked by the 2018 Times Higher Education ranking of U.S. Universities (bottom to top).
		Points correspond to organizations, and a sample of individual states are highlighted.
		Full organization names listed in Table~\ref{table:supp:orglabels}.
		\textbf{b.} Comparison between the ranking of organizations by their rank in the Times ranking and the SemAxis rank derived from the similarity to the \textit{Elite} end of the axis prestige axis \textit{a}.
		White points are those top and bottom 20 universities aggregated to define the ends of the axis.
		The grey box corresponds to the top 20 and bottom 20 ranks.
	  	The estimate of the spearman rank correlation between the SemAxis and Times ranking is 0.81 (n = 145).
	}
	\label{fig:semaxis}
\end{figure}

@@

where $m_{i}$ and $m_{j}$ are the yearly-average size of each organization during the time window, and $f(r_{ij})$ is a distance-decaying function  between the two organizations. There are two popular forms of the $f(r_{ij})$. The first one is power-law function with $f(r_{ij})= r^{-\alpha}_{ij}  (\alpha > 0)$, and the another one is exponential function with $f(r_{ij}) = e^{-\beta r_{ij}} (\beta > 0)$ \autocite{chen2015distance}.  $f(r_{ij})$, $C$ can be chosen as to fit the data with the simple regression.

@@

%
% SUBSECTION: SEMAXIS
%
\subsection*{SemAxis}
By constructing semantic embeddings from mobility trajectories, we allow for unique methodology for understanding the embedding space, such as SemAxis.
SemAxis~\autocite{an2018semaxis} is a technique that leverages the latent semantic characteristics of word embeddings to represent the position of terms along a conceptual axis, reflecting the relationship of these term to the concept.
The precise axis is an antonym pair defined \textit{ad-hoc} for the specific use case.
In the case of language, the ends of the axis could be ``good'' and ``bad'', ``surprising'' and ``unsurprising'', or ``masculine'' and ``feminine''. It is shown that projection of words to given specific axis fit well with widely shared cultural connotation from surveyed
responses and labeled historical data \autocite{kozlowski2018geometry}.
In the case of scientific mobility, SemAxis offers a means of operationalizing abstract relationships between organizations.

>>>>

As scientist's careers progress, they move between organizations or pick up additional affiliations, forming co-affiliation or \textit{mobility trajectories} that can be derived from their publication histories~\autocite{robinson2019mobility, sugimoto2017mostimpact}.
By treating geo-located organizations analogously to words and trajectories analogously to sentences, we learn vector-space embedding of organizations using a neural embedding method (Fig.~\ref{fig:methods:mobility_sentence}).
In the resulting embedding, each organization corresponds to a vector, and vectors relate to one another \textit{semantically}--just as with language, organizations are \textit{semantically} similar when they share similar co-affiliation \textit{contexts}. 
We measure the semantic distance between organization vectors using \textit{cosine distance}, and use this to model distance in the landscape of scientific mobility.

@@

Where the \textit{expected flux}, $\hat{T_{ij}}$, or expected number of authors co-affiliated with a pair organizations during the time period is proportional to the organization's sizes (both mobile and non-mobile authors, see Data and Methods), $m_{i}$ and $m_{j}$, and decays as a function of their distance, $f(r_{ij})$.
We consider two distances, the geographic distance between the organizations as a baseline, and the \textit{embedding distance} defined as the cosine distance between the organization's vector in the trained embedding space.
We argue that the performance of a distance measure can be observed from their ability to explain expected mobility between organizations and to make predictions using the gravity model.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

Given that embeddings capture effective distance between organizations, we can now examine the global structure of scientific mobility.
Because the embeddings are defined in a high-dimensional space, we apply a widely-used dimensionality reduction technique (UMAP~\autocite{mcinnes2018umap}) to visualize the overall structure (Fig.~\ref{fig:projection}a).
Globally, the effect of geography is conspicuous; organizations tend to form clusters based on their national affiliations and national clusters tend to be near their geographic neighbors. 
At the same time, the embedding space also reflects a mix of geographic, historic, cultural, and linguistic relationships between regions much more clearly than alternative network visualizations (Fig.~\ref{fig:supp:network_vis}).

@@

The results demonstrating embeddings' ability to capture various features governing mobility raise the question of whether they can also capture semantic axes analogously to word embeddings. 
For instance, does the mobility embedding encode information about the prestige or universities?
We show that, in addition to encoding national characteristics, the embedding also encodes domestic geography and latent hierarchy.

By treating organizations analogously to words and trajectories to sentences, we embed organizations in a way that preserves \textit{semantic} relationships, meaning organizations that share similar co-affiliation \textit{contexts} will in turn be similar (see Data and Methods). 
Leveraging these unique semantic properties, we employ the SemAxis approach~\autocite{an2018semaxis} and define two axes.
The first is the California--Massachusetts axis defined by the mean vector of universities in California on one end, and in Massachusetts on the other. 
The second axis is the low- to high- prestige axis;
the elite end of the axis is defined as the top 20 U.S. universities according to the Times Ranking of World Universities, whereas the non-elite axis is defined as bottom-ranked and geographically matched (by U.S. census region) set of universities.
A university projected onto the axis corresponds to their relative nearness to the mean vector defining each pole (Fig.~\ref{fig:semaxis}a).

Generally, universities in Western states, such as Arizona (Fig.~\ref{fig:supp:semaxis_states}b) are nearer to the California end of the axis, whereas states in the Northeast such as New York (Fig.~\ref{fig:supp:semaxis_states}f) and Connecticut (Fig.~\ref{fig:supp:semaxis_states}e) are nearer the Massachusetts end.
Universities geographically distant from both California and Massachusetts are less aligned with either, such as Texas (Fig.~\ref{fig:supp:semaxis_states}g)s and Florida (Fig.~\ref{fig:supp:semaxis_states}i).

Universities hiring and mobility networks are known to exhbit latent pretige hierarchies ~\autocite{deville2014career, clauset2015hierarchy} that should be encoded in the embedding space.
Accordingly, prestigious universities such as Stanford, Colombia, MIT, Harvard, and NYU, group towards the top of the axis.
Universities at the bottom of this axis tend to be regional universities with lower national profiles.
By considering mobility between all organization types, organization embeddings offer a new means of reasoning about the prestige of organizations typically excluded from rankings such as government organizatioss (Fig.~\ref{fig:supp:semaxis_sectors}a), research institutes (Fig.~\ref{fig:supp:semaxis_sectors}b), and teaching colleges (Fig.~\ref{fig:supp:semaxis_sectors}c)
\ref{fig:supp:sem

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_fig.pdf}
	\caption{
		\textbf{Embedding captures latent geography and prestige hierarchy.}
		\textbf{a.} illustrates SemAxis projection along two axes, comparing California to Massachusetts universities (left to right), and between the top 20 and geographically-matched bottom 20 universities ranked by the 2018 Times Higher Education ranking of U.S. Universities (bottom to top).
		Points correspond to organizations, and a sample of individual states are highlighted.
		Full organization names listed in Table~\ref{table:supp:orglabels}.
		\textbf{b.} Comparison between the ranking of organizations by their rank in the Times ranking and the SemAxis rank derived from the similarity to the \textit{Elite} end of the axis prestige axis \textit{a}.
		White points are those top and bottom 5 universities aggregated to define the ends of the axis.
		The grey box corresponds to the top 5 and bottom 5 ranks.
	  		  	Even when considering only a total of ten organization vectors, the estimate of the spearman rank correlation between the SemAxis and Times ranking is 0.73 (n = 145).
	}
	\label{fig:semaxis}
\end{figure}

@@

where $m_{i}$ and $m_{j}$ are the organization masses, or the yearly-average count of unique authors affiliated with each organization during the time window, and $f(r_{ij})$ is a distance-decaying function  between the two organizations. There are two popular forms of the $f(r_{ij})$. The first one is power-law function with $f(r_{ij})= r^{-\alpha}_{ij}  (\alpha > 0)$, and the another one is exponential function with $f(r_{ij}) = e^{-\beta r_{ij}} (\beta > 0)$ \autocite{chen2015distance}.  $f(r_{ij})$, $C$ can be chosen as to fit the data with the simple regression.

@@

%
% SUBSECTION: SEMAXIS
%
\subsection*{SemAxis}
By constructing semantic embeddings from mobility trajectories, we allow for unique methodology for understanding the embedding space, such as SemAxis.
SemAxis~\autocite{an2018semaxis} is a technique that leverages the latent semantic characteristics of word embeddings to represent the position of terms along a conceptual axis, reflecting the relationship of these term to the concept.
The precise axis is an antonym pair defined \textit{ad-hoc} for the specific use case.
In the case of language, the ends of the axis could be ``good'' and ``bad'', ``surprising'' and ``unsurprising'', or ``masculine'' and ``feminine''. It is shown that projection of words to given specific axis fit well with widely shared cultural connotation from surveyed
responses and labeled historical data \autocite{kozlowski2018geometry}.

In this analysis, we embed organization identifiers and trajectories, rather than words and sentences, however mobility can be thought of semantically, just like language. 
Under this view, organizations can be thought of as similar when they have similar \textit{contexts}--that is, authors affiliated with that organization also tend to co-affiliate the same set of other organizations.
By treating organizations analogously to words and trajectories analogously to sentences, we can use SemAxis to leverage the semantic properties of the resulting embedding vectors to operationalize abstract relationships between organizations.
